\documentclass[a4paper, oneside, 10pt, final]{memoir}
\input{packages}
\chapterstyle{culver}

\usepackage{fixme}
\usepackage[english]{babel}
%\usepackage[utopia]{mathdesign}

\usepackage[osf,sc]{mathpazo}
\linespread{1.05}
%\usepackage{fourier}

\usepackage{semantic}
\author{Jesper Louis
  Andersen\\jesper.louis.andersen@gmail.com\\140280-2029}
\title{Formalizing a Virtual Machine}
\date{\today}

\newlength{\drop}
\newcommand*{\titleM}{\begingroup% Misericords, T&H p 153
  \drop = 0.08\textheight
  \centering
  {\Huge\bfseries Lambda}\\[\baselineskip]
  {\scshape IR of exsml}\\[\baselineskip]
  {\scshape by}\\[\baselineskip]
  {\large\scshape Jesper Louis Andersen\\jesper.louis.andersen@gmail.com}\par
  \endgroup}

\bibliographystyle{plain}

\begin{document}
\maketitle{}
\chapter*{Problem statement}

\emph{Virtual Machines} (VMs) is a popular way to achieve hardware
independence: one writes a compiler targetting the VM and then
leverages on it to provide backends for different hardware
architectures. There are several virtual machines, which are
extensively used in todays computing-infrastructure. Prime examples
are the Java JVM and the .NET framework CLR.

Both the JVM and CLR provides a high-level approach to the VM: the VM
operations translate to several machine code instructions. Another
approach is to provide a low-level VM where each instruction maps to
few machine instructions. The Low-level Virtual Machine (LLVM) is one
such machine.

LLVM provides a VM-language called the LLVM IR. This language is,
essentially, a typed assembly language with an
SSA-form\cite{appel:1998:modern} requirement.

However, the LLVM IR language has no formal semantics. It is specified by an
informal text-description. In addition, it is not clear that the type
system for the IR has the property of type safety\cite{pierce:2002:types}.

Our thesis is:
\begin{center}
  ``It is feasible and useful to produce a formal semantics and type
  system for a subset of the LLVM IR''
\end{center}
Our formalization will probably digress and not follow the LLVM IR
blindly. Rather, we will let us inspire by the constructions in the IR
but diverge where formalization will be too hard. Also we stress that
we will analyze a subset of the IR. The full IR is too time-consuming
to formalize and we do not know if the type system is type safe. We
aim to produce an operational semantics which \emph{is} type-safe,
even if this fragment is rather small.

An operational semantics of the fragment is useful in the sense that
formalizations often gives deeper insight than informal
descriptions. We hope to gain some additional insights in the
formalization work.

To embrace the state-of-the-art of formalization, we plan we verify
correctness by translating the semantics, and if time allows key
proofs, into the Twelf Logical Framework. This will enable us to
verify correctness by machine.

At the project end, we will have produced:
\begin{itemize}
\item A typed assembly language (TAL), inspired by the rules of the LLVM
  IR. This TAL will be given an operational semantics.
\item An encoding of key parts of the semantics and its properties in
  Twelf; if time allows, everything.
\item A report describing the product.
\end{itemize}

\paragraph{Related work}

Typed assembly languages have seen much work in the litterature, see
for instance \cite{crary:2003:toward},
\cite{morrisett.crary.ea:1999:talx86}. There are even formalizations
of some of these languages in Twelf. This allows us to use earlier
experiences in working with assembly-style language. To our knowledge
however, there has been no attempts at formalizing LLVM in Twelf.

\chapter*{Learning objectives}

\begin{itemize}
\item Our primary objective is to become better at formalizing informal
  descriptions of semantics and type systems into inference rules.
\item Second, we will hone our skills on showing key properties of
  formalized systems by proof.
\item Finally, our objective is to verify the correctness of our
  formalization by transscribing it into the Twelf Logical Framework.
\end{itemize}

\bibliography{biblio}
% State what the goals of this project is.
% State we focus on the operational semantics.
% State we will use Twelf where applicable if time allows.

\end{document}
